---
title: "Modeling Choices #02. T-distribution"
author: 
  - name: Joon-Ho Lee (joonho@berkeley.edu)
date: "February 23, 2020"
output:
  html_document: 
    css: styles.css
    fig_caption: yes
    highlight: haddock
    number_sections: yes
    theme: readable
    toc: yes
    toc_depth: 2
  tufte::tufte_html:
    number_sections: yes
    toc: true
    toc_depth: 2
  pdf_document:
    toc: yes
    toc_depth: '3'
bibliography: bibliography.bib
---

```{r basic_setup, include=FALSE}
### Set working directory
setwd("~/Treatment-effect-heterogeneity/Multisite Trials/docs/02_Modeling choices_02_T5")

### Set RMarkdown options
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE)

### Call libraries
library(tidyverse)
library(cowplot)
library(hhsim)
library(rstan)
library(bayesplot)

### Set Stan options
# options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
Sys.setenv(LOCAL_CPPFLAGS = '-march=native')

### Theme settings
theme_preset <- 
  theme_bw() + 
  theme(panel.background = element_blank(),
        panel.grid = element_blank(), 
        legend.position = "bottom", 
        legend.direction = "horizontal", 
        legend.title = element_blank())
```


# The @rubin1981estimation model with $t$ shrinkage prior 

Consider $K$ study sites in which researchers perform the same interventions and measure the same outcomes. Each site, indexed by $k$, estimates a treatment effect $\tau_k$ averaged across individuals in the site. The sites don't report $\{\tau_k\}^{K}_{k=1}$: instead, they report $\{\hat{\tau}_k\}^{K}_{k=1}$. Some of the observed variation in $\{\hat{\tau}_k\}^{K}_{k=1}$ is sampling variation, yet there is likely to be some genuine variation in effects across settings, often defined as $\sigma_{\tau}^{2} = \text{var}(\tau_{k})$. 


@rubin1981estimation considers a case in which the analyst has access to a set of estimated effects $\{\hat{\tau}_k\}^{K}_{k=1}$ and estimates of the associated sampling errors $\{\hat{se}_{\tau_k}\}^{K}_{k=1}$. Rubin specifies a relationship between the observed estimates and the unobserved $\{\tau_k\}^{K}_{k=1}$, and in addition spcifies a relationship between $\{\tau_k\}^{K}_{k=1}$ and the aggregate parameters $(\tau, \sigma_{\tau}^2)$). The @rubin1981estimation model has a hierarchical likelihood in which each site has own treatment effect parameter, $\tau_k$, but these effects are all drawn from a common distribution governed by $(\tau, \sigma_{\tau}^2)$ as follows:

$$
\hat{\tau}_{k} \sim N(\tau_{k}, \hat{se}_{k}^2) \quad \forall k, 
$$
$$
\tau_{k} \sim t_{\nu = 5}(\tau, \sigma_{\tau}^2) \quad \forall k,  
$$

where $\mathbb{E}(\tau_k) = \tau$ and $\text{var}(\tau_k) = \sigma_{\tau}^{2} \cdot \frac{\nu}{\nu - 2}$. A $t$ prior for a normal mean provides a reasonable posterior even if the data and prior disagree. A $t$ prior provides similar shrinkage to a normal prior when the data and prior agree, but provides little shrinkage when the data and prior disagree. By using the $t$ shrinkage prior, we expect to go beyond normality to add heavy tails. 



# Generate simulated data from a $t_{\nu = 5}$ distribution 

$G$ will be simulated to follow a student $t$ distribution with degrees of freedom $\nu = 5$,  mean 0, and variance 1. We first set our simulation parameters to reflect a range of data informativeness and heterogeneity among sites with respect to variance. 


```{r, fig.height=4, fig.width=5}
### Example settings
n <- 50; rgt <- 1; rsr <- 5

### Set maximum and minimum within-site variances
sigma2max <- rgt*rsr
sigma2min <- sigma2max/(rsr^2)

### Generate within-site variances (sds) for K sites
sigma2 <- exp(seq(from = log(sigma2min), to = log(sigma2max), length = n))
sd <- sqrt(sigma2)

### Plot within-site variances
df_temp <- tibble(sigma2, sd) %>%
  mutate(K = 1:n())

ggplot(data = df_temp, aes(x = K, y = sigma2)) + 
  geom_point(size = 2.0) + geom_path(size = 1.0) + 
  scale_x_continuous(breaks = seq(1, n, by = 5)) + 
  labs(title = bquote("The heterogeneity of the " ~ sigma^2 ~"s")) + 
  theme_preset
```


The **R** function that follows simulates a dataset conforming to the $t_{\nu = 5}$ model.

```{r}
### Define a function to generate random draws from T-distribution with df = 5 (nu)
GenerateT <- function(n, rgt, rsr, nu) {
  
  ### Generate within-site variances (sds) for K sites
  sigma2max <- rgt*rsr
  sigma2min <- sigma2max/(rsr^2)
  sigma2 <- exp(seq(from = log(sigma2min), to = log(sigma2max), length = n))
  sd <- sqrt(sigma2)
  
  ## Mean of T-distribution
  mn <- 0
  
  ## Generate data object
  data_obj <- list()
  data_obj$tau_k <- rt(n, nu)*sqrt((nu - 2)/nu)
  data_obj$tau_k_hat <- rnorm(n, data_obj$tau_k, sd)
  data_obj$se_k2 <- sigma2
  return(data_obj)
}

### Generate sample draws from the T model
df_G <- GenerateT(n = 50, rgt = 1, rsr = 5, nu = 5) %>%
  data.frame() %>%
  rownames_to_column("K") %>%
  mutate_at(.vars = c("K"), .funs = as.numeric)

head(df_G)
```


Next, we compare the observed estimates ($\hat{\tau}_k$) to the true distribution of $\tau_k$.  


```{r, fig.width=10}
### Define a function to generate plot
plot_compare_true_obs <- function(df_G){
  
  # Compare the true distribution (tau_k) to the observed data (tau_k_hat)
  df_plot <- df_G %>%
    select(tau_k, tau_k_hat) %>% 
    rename(True = tau_k, Observed = tau_k_hat) %>%
    gather(key = variable, value = value)
  
  p1 <- ggplot(data = df_plot, aes(x = value, group = variable, fill = variable)) +
      geom_density(position = "identity", size = 0.1, alpha = 0.3) + 
      geom_vline(aes(xintercept = 0), size = 0.3, color = "red", linetype = "dashed") + 
      labs(title = bquote("True vs. Observed " ~ tau_k ~" Distributions"), 
           x = expression(tau_k)) + theme_preset

  # Compare point estimates by sigma_k
  p2 <- ggplot(data = df_G %>% mutate(gap = abs(tau_k_hat - tau_k)), 
               aes(x = se_k2, y = gap)) + 
      geom_point(aes(size = se_k2)) + 
    labs(title = bquote("Gap between True and Observed tau_k vs. " ~ se^2), 
         y = bquote("Gap between True and Observed tau_k"), 
         x = expression(se^2)) + theme_preset
  
  # Combine the two plots
  p_grid <- plot_grid(p1, p2, labels = "AUTO")
  return(p_grid)
}

plot_compare_true_obs(df_G)
```


The empirical distribution function (EDF) of the observed $\hat{\tau}_k$ is overdispersed due to the uncertainty of the observed data, $\hat{se}_k^2$. The discrepancy between true and observed $\tau_k$s tends to be larger for the sites with larger $\hat{se}_k^2$.   



# Parameter estimation using **Stan**

First, we obtain simulates from the posterior of a $t_{\nu = 5}$ model using **Stan**. The **Stan** program for the @rubin1981estimation model with $t$ shrinkage prior is provided below:


```
data {
  int<lower = 0> K;         // number of sites 
  real tau_hat_k[K];        // estimated treatment effects
  real<lower=0> se_k[K];    // s.e. of effect estimates 
}
parameters {
  real tau; 
  real<lower=0> sigma_tau;
  real tau_k[K];
}
model {
  sigma_tau ~ uniform(0, 1000);           // vague prior
  tau ~ normal(0, 1000);                  // vague prior
  tau_k ~ student_t(5, tau, sigma_tau);   // second-level normal
  tau_hat_k ~ normal(tau_k, se_k);        // third-level normal
}
generated quantities{
  real predicted_tau_k;
  predicted_tau_k = student_t_rng(5, tau, sigma_tau);  //mixed predictive 
}
```

Note that the Rubin model assumes $t_{\nu = 5}$ distribution of effects across sites. The Rubin model performs univariate shrinkage so we will not need to specify a hyper-correlation prior. But we still need to specify hyperpriors on the hyper-mean $\tau$ and hyper-SD $\sigma_{\tau}$:

- For $\tau$, the prior is $\text{N}(0, 1000)$. This is a very weak prior which does little regularization, centered at zero. It assumes that causal effects should not be thought of as large unless data contains evidence to the contrary. 

- For $\sigma_{\tau}$, the prior is $\text{Uniform}(0, 1000)$, which is also a very weak prior. 


The model specified in the Stan program is fit to the simulated data:

```{r, warning=FALSE, results=FALSE}
### Collect data into a list format suitable for Stan
stan_data <- list(K = length(df_G$K), 
                  tau_hat_k = df_G$tau_k_hat, 
                  se_k = sqrt(df_G$se_k2))

### Compile and run the stan model
fit_rubin<- stan(file = "rubin_model_02_T5.stan",
                 data = stan_data,
                 iter = 1000, chains = 4)
```


The posterior means and standard deviations of the parameters can be displayed as follows:

```{r}
print(fit_rubin, probs = c(0.1, 0.5, 0.9), digits = 3)
```


Before interpreting the results, it is necessary to check that the chains have converged for each parameter. The $\hat{R}$ statistics shown in the rightmost column of the model summary are all less than 1.1. Also the effective sample size estimates are sufficient for inference. Thus it seems that **Stan** has produced an adequate approximation of the posterior.

Then, we extract posterior samples and save it for later comparison with estimates from `hhsim` package. 

```{r}
### Extract posterior samples
posterior <- as.data.frame(fit_rubin) %>%
  mutate(Type = "Stan") %>%
  dplyr::select(Type, everything())
dim(posterior)
```




# Parameter estimation using the `hhsim` package

Second, we obtain simulates from the posterior of a $t_{\nu = 5}$ model using the `hhsim` package. The **Stan** program above runs 4 chains with a burn-in period of 500 iterations followed by 500 additional iterations per chain. Since the `hhsim` package does not use separate chains, we set the number of  MCMC and burn-in iterations to 2,000, respectively, in order for the total iterations to be the same with those of the **Stan** program.    


```{r}
### Set numbers of iterations
nmcmc <- 2000    # number of MCMC iterations per each model fit
nburn <- 2000    # burn-in iterations to discard at each round
nmcmcall <- nmcmc + nburn  

### Extract vectors of observed tau_ks and their SEs
Y <- df_G$tau_k_hat   # a vector of observed tau_ks
sigma2 <- df_G$se_k2  # a vector of known SEs
nDraws <- nmcmcall    # number of MCMC draws

### Estimate the T5 model using hhsim package
outp <- GaussianT(Y, sigma2, nDraws, nu = 5)

### Define a function to tidy up the output object
get_posterior_sample <- function(output = outp, nburn = 100, nmcmcall = 600){
  
  # (1) Tidy up the "theta" output object
  df_theta <- outp[["theta"]] %>%
    unlist() %>%
    matrix(nrow = length(outp[["theta"]]), byrow = TRUE) %>%
    data.frame() %>%
    slice(c((nburn + 1):nmcmcall))  # throw away burn-ins
  colnames(df_theta) <- paste0("tau_k[", seq(ncol(df_theta)), "]")
  
  # (2) Tidy up the "tau" and "sigma_tau" output objects
  df_hyperparm <- outp[["m"]] %>%
    cbind.data.frame(outp[["tau2"]]) %>%
    slice(c((nburn + 1):nmcmcall))  # throw away burn-ins
  colnames(df_hyperparm) <- c("tau", "sigma_tau")
  
  ### Return the resulting object
  cbind.data.frame(df_hyperparm, df_theta)
}

### Save posterior samples
posterior_hhsim <- get_posterior_sample(outp, nburn, nmcmcall) %>%
  mutate(Type = "hhsim") %>% dplyr::select(Type, everything())
```



# Comparison of posterior samples obtained from **Stan** and `hhsim`

Would the **Stan** program and `hhsim` package lead to the same estimates? We start from combining the two data frames containing posterior samples:

```{r}
### Combine two dataframes: Stan and hhsim
df_combine <- posterior %>%
  dplyr::select(-predicted_tau_k, -lp__) %>% 
  bind_rows(posterior_hhsim) %>%
  mutate(Type = factor(Type, levels = c("Stan", "hhsim")))
```


From the results presented below, we conclude that the **Stan** program and the `hhsim` package generates more or less similar posterior distributions for (1) hyperparameters ($\tau$ and $\sigma_{\tau}$), (2) empirical distribution function (EDF) estimates for $\tau_{k}$s, and (3) site-specific individual effect estimates $\tau_{k}$s.     


## Hyperparameters $\tau$ and $\sigma_{\tau}$

```{r, fig.width=10}
### Generate a dataframe to plot
df_plot <- df_combine %>%
  dplyr::select(Type, tau, sigma_tau) %>%
  gather(key = variable, value = value, -Type) %>%
  mutate(variable = factor(variable, levels = c("tau", "sigma_tau")))

### Define a function to generate plot
plot_compare_density <- function(df_plot, title = title){
  ggplot(data = df_plot, aes(x = value, group = Type, fill = Type)) +
    geom_density(position = "identity", size = 0.1, alpha = 0.3) + 
    geom_vline(aes(xintercept = 0), size = 0.3, color = "red", linetype = "dashed") + 
    labs(title = title) + theme_preset
}

plot_compare_density(df_plot, title = "Posterior densities of hyperparameters (Stan vs. hhsim)") + 
  facet_wrap(~ variable, scales = "free")
```


## Empirical distribution function (EDF) estimates for $\tau_{k}$s 

```{r}
df_temp <- df_combine %>%
  dplyr::select(Type, contains("tau_k[")) %>%
  gather(key = variable, value = value, -Type) %>%
  group_by(Type, variable) %>%
  summarise(mean = mean(value, na.rm = TRUE), 
            sd = sd(value, na.rm = TRUE)) %>%
  mutate(id = as.numeric(str_extract(variable, "\\d+"))) %>%
  arrange(id, Type)

head(df_temp, 10)  
```


```{r, fig.width=10}
p1 <- plot_compare_density(df_temp %>% rename(value = mean), title = "EDF of posterior means")
p2 <- plot_compare_density(df_temp %>% rename(value = sd), title = "EDF of posterior SDs")
plot_grid(p1, p2, labels = "AUTO")
```



## Site-specific individual effect estimates $\tau_{k}$s


```{r, fig.width=10, fig.height=7}
### Compare densities: site-specific estiamtes
set.seed(12345)
rand_var <- paste0("tau_k[", sample(1:50, 10, replace = FALSE), "]")

df_plot <- df_combine %>%
  dplyr::select(Type, rand_var) %>%
  gather(key = variable, value = value, -Type)

title <- "Posterior densities of site-specific estimates (Stan vs. hhsim)"
plot_compare_density(df_plot, title = title) + 
  facet_wrap(~ variable, scales = "free")
```


# References



