---
title: "Triple-goal estimates in multilevel models"
author: 
  - name: Joon-Ho Lee (joonho@berkeley.edu)
date: "February 3, 2020"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: yes
    theme: journal
    highlight: haddock
    css: styles.css
    fig_caption: yes
  tufte::tufte_html:
    number_sections: yes
    toc: true
    toc_depth: 2
  pdf_document:
    toc: yes
    toc_depth: '3'
bibliography: bibliography.bib
---

```{r basic_setup, include=FALSE}
# Set working directory
setwd("~/Treatment-effect-heterogeneity/Multisite Trials/docs/Triple-goal estimator")

# Set RMarkdown options
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE)

# # Set Stan options
# options(mc.cores = parallel::detectCores())
# rstan_options(auto_write = TRUE)
# Sys.setenv(LOCAL_CPPFLAGS = '-march=native')

# Call libraries
library(tidyverse)
library(HETOP)
```

# Introduction

@shen1998triple developed a new method of parameter estimation in two-stage hierarchical models. 
 



# Generating a simulated data

```{r}
# Simulate the true distribution of theta
set.seed(1001)
K <- 50           # Number of sites
nsamp <- 500      # Number of posterior samples
theta_true <- rnorm(K, mean = 0, sd = 1)


# Plot the true distribution
df <- theta_true %>% tibble()
names(df) <- "theta_true"
p <- ggplot(data = df, aes(x = theta_true)) + 
  geom_histogram() +
  theme_minimal() + 
  labs(title = "The distribution of true theta")
print(p)
```




```{r}
# Simulate the posterior samples of each theata_k
s_fix <- matrix(theta_true, ncol = K, nrow = nsamp, byrow = TRUE)
s_err <- matrix(rnorm(K*nsamp, sd = 0.4), ncol = K, nrow = nsamp) 
s_mat <- s_fix + s_err
dim(s_mat)
```



# Posterior means

```{r}
# Get posterior means
theta_pm <- apply(s_mat, 2, mean)  # Posterior means
etadot <- mean(theta_pm)           # Mean of the posterior means
var_pm <- var(theta_pm)            # Variance of the posterior means


# Get posterior SDs and variances
theta_psd <- apply(s_mat, 2, sd)   # Posterior SDs
lambda_k <- theta_psd^2            # Posterior variances
```




# Contrained Bayes estimates

```{r}
# Get constrained Bayes (CB) estimates
theta_cb <- etadot + (sqrt(1 + (mean(lambda_k)/var_pm))*(theta_pm - etadot))
```





# Triple-goal estimates


```{r}
# (1) Non-integer ranks
#  Get rowwise ranks over 500 posterior samples and get averages of them
rbar <- apply(t(apply(s_mat, 1, rank, ties.method = "average")), 2, mean)
print(rbar)

# (2) Integer ranks (rank estiamte)
rhat <- rank(rbar, ties.method = "random")
print(rhat)
```




```{r}
# Triple-goal (GR) estimates
quantile.type <- 7

theta_gr <- quantile(c(s_mat), probs = (2 * rhat - 1)/(2 * K), 
                     names = FALSE, type = quantile.type)
```




# Compare distributions {.tabset .tabset-fade .tabset-pills}

```{r}
# Collect all estimates
df_est <- data.frame(index = 1:K, theta_true = theta_true, 
                     theta_pm = theta_pm, theta_cb = theta_cb, theta_gr = theta_gr, 
                     theta_psd = theta_psd, rbar = rbar, rhat = rhat)

round(df_est, 2)
```



```{r}
# Select only theta estimates and convert to a long format
df_est_long <- df_est %>%
  dplyr::select(contains("theta")) %>%
    dplyr::select(-contains("psd")) %>%
  gather(key = estimator, value = value)


# Generate a function to compare two distributions
plot_compare_dist <- function(data = df_plot, vec_est = vec_est){
  
  df_plot <- df_est_long %>% 
    filter(estimator %in% vec_est)
  
  ggplot(df_plot, aes(x = value, group = estimator, color = estimator)) + 
    geom_density(position = "identity", size = 1) + 
    theme_classic()
}
```

- Click tabs

## True vs. Posterior means

```{r}
plot_compare_dist(df_plot, vec_est = c("theta_true", "theta_pm"))
```


## True vs. Constrained Bayes estimates

```{r}
plot_compare_dist(df_plot, vec_est = c("theta_true", "theta_cb"))
```


## True vs. Triple-goal estimates

```{r}
plot_compare_dist(df_plot, vec_est = c("theta_true", "theta_gr"))
```





# References



